Extended Data

### 2021-06-19 doc/manuscript/Dow_manuscript_Nature_cd.Rmd
some work on draft

They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.* (Add a line suggeesting this may be because of water stress?)
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
*It would be interesting to compare the magnitude of these advances (days per degree C)*
However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possibile mechanisms.
First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
Alternatively, warm springs may be associated with greater summer drought stress... [@buermann_widespread_2018]

@@

Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
We show that the extra C does not go into woody growth, raising the question of where this C goes.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013]. 
Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].
*(we can easily test this: [GitHub issue 57](https://github.com/EcoClimLab/growth_phenology/issues/57))*
Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002]. 
Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome  showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.

>>>>

They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance. A larger build-up of drought stress due to earlier leaf up increasing seasonal water demand may prevent extra C from being stored in woody biomass. (Add a line suggeesting this may be because of water stress?)
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the phenology of tree stem growth but have little effect on annual woody productivity (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in the timing of stem growth milestones under warmer springs is consistent with similar phenological advances observed for leaf development (Fig. `r DOY_timing`)[@keenan_net_2014] and NEE [@keenan_net_2014].
*It would be interesting to compare the magnitude of these advances (days per degree C)*
However, inconsistent with leaf phenology and NEE, we observed that the length of the period of most active stem growth does not increase under warmer springs. 
This suggests that the growth benefit of warmer springs is compensated by negative effects later in the growing season, with two primary possibile mechanisms.
First, stem growth, like leaf phenology[@zani_increased_2020], may be sink-limited, such that an advance in the start of growth ($DOY_25$) accompanied by little change in growth rate ($g_{max}$) results in an earlier end to the season of rapid growth ($DOY_{75}$).
Alternatively, warm springs may be associated with greater seasonal water stress, effectively canceling out any positive effects of an extended growing period.  [@buermann_widespread_2018]

@@

Our finding that interannual variation in woody growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with parallel findings for NEE [@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect between the woody growth responses to spring temperatures observed here and parallel studies on NEE implying that annual C sequestration increases -- at least modestly -- with spring temperatures[@keenan_net_2014] or with the length of the carbon uptake period[@zhou_explaining_2016; @fu_maximum_2019].
We show that the extra C does not go into woody growth, raising the question of where this C goes.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.
First, C that is fixed in late summer or fall may be used for woody growth that is not apparent from diameter measurements in the current year: either used for thickening of cell walls, a process that lags behind stem expansion[@cuny_woody_2015], or saved within trees as non-structural carbohydrates and used towards growth the following year[@trumbore_quantifying_2002;@richardson_seasonal_2013].(Could it also be used to repair drought-related damage in trees? Like the cavitation of xylem in ring-porous species?) 
Tree-ring evidence from SCBI suggests that both processes may influence the climate sensitivity of annual woody growth, but that this dynamic is complex and unlikely to result in substantial additional growth in the following year [@helcoski_growing_2019].
*(we can easily test this: [GitHub issue 57](https://github.com/EcoClimLab/growth_phenology/issues/57))*
Second, any additional C that is not rapidly respired may be allocated, likely in the following growing season, to plant functions other than stem growth, including respiration, reproduction, foliage, fine roots, or root exudates[@trumbore_quantifying_2002]. 
Most of this carbon would have a relatively low residence time in the ecosystem[@trumbore_quantifying_2002], although a small fraction would be incorporated into soil organic carbon pools *with residence times rivaling that of wood [REFS]*.
Notably, both of these involve lags of a year or more and would not be captured in interannual variation in NEE, implying that the long-term effect of lengthened growing season on NEE may not be captured in studies, including this one, that consider time frames of a year or less[@zhou_explaining_2016; @fu_maximum_2019].
Studies examining long-term trends in growing season length and ecosystem C uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014] -- as opposed to their interannual variation -- within or including the temperate deciduous biome  showed increasing trends in both variables, suggesting that the C not allocated to current year growth may have a fairly long residence time within the ecosystem; however, the amount and fate of this carbon remains unknown.
